Optimizing Abstract Abstract Machines

### 2012-11-12 paper/paper.tex

%% \paragraph{Primitives}

%% \begin{align*}
%% \mnum+1 &\in \interpdelta(\saddone,\mnum) &
%% \mnum-1 &\in \interpdelta(\ssubone,\mnum)\\
%% \strue &\in \interpdelta(\szerohuh,\szero) &
%% \sfalse &\in \interpdelta(\szerohuh,\mnum)\text{ if }\mnum\neq \szero\\
%% \end{align*}

%% \begin{align*}
%% \sNum &\in \hat\interpdelta(\saddone,\mnum) &
%% \sNum &\in \hat\interpdelta(\ssubone,\mnum)\\
%% \strue &\in \hat\interpdelta(\szerohuh,\sNum) &
%% \sfalse &\in \hat\interpdelta(\szerohuh,\sNum)\\
%% \strue &\in \hat\interpdelta(\szerohuh,\szero) &
%% \sfalse &\in \hat\interpdelta(\szerohuh,\mnum)\text{ if }\mnum\neq \szero\\
%% \end{align*}

%\newpage
\section{Baseline analyzer}
\label{sec:baseline}

@@

\section{Mostly precision preserving recipe}
\label{sec:opt}

@@

\begin{itemize}
 \item{[SS] quotient by sets of values (lazy non-determinism)}
 \item{[SS] remove {\tt ev} states (abstract compilation)}
 \item{[DS] operate on store changes rather than whole stores (store deltas)}
 \item{[DS] represent the store as a global, mutable hash and
   determine a state's store by the number of changes made to the
   global store at the time of visiting (timestamp)}
 \item{[DS] use a global mutable workset}
 \item{[DS] represent the global store as a global, mutable vector in
   the 0-CFA case. (preallocation)}
\end{itemize}

Each technique requires minimal changes to the reduction
relation/fixpoint, and each improves performance by integer
factors. The first three remain in the purely functional realm. The
final three use mutation in small, localized places, so they do not
increase conceptual complexity drastically.

@@

All following techniques preserve precision.
% Lazy:  cpu time: 32481 real time: 32881 gc time: 547

@@

Notice that the only difference between concrete and abstract interpretation
is which addresses are pushed.

\begin{figure}
\begin{align*}
\compile{\svar\mvar} &= \lambda(\menv,\msto,\mkont) .\co{\mkont,\mval,\msto} \text{ where }\mval\in\msto(\menv(\mvar))
\\
\compile{\slit\mlit} &= \lambda(\menv,\msto,\mkont) .
\co{\mkont,\mlit,\msto}
\\
\compile{\slam\mvar\mexp} &= \lambda(\menv,\msto,\mkont) .
\co{\mkont,\clos{\mvar,\compile\mexp,\menv},\msto}
\\
\compile{\sapp[^\mlab]{\mexpi0}{\mexpi1}} &= \lambda^\mcntr(\menv,\msto,\mkont) .
\compile{\mexpi0}^\mcntr(\menv,\msto',\kar[_\mlab^\mcntr]{\compile{\mexpi1},\menv,\maddr})
\\
&
\text{ where }\maddr,\msto' = \mathit{push}^\mcntr_\mlab(\msto,\mkont)
\\
\compile{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2}} &= \lambda^\mcntr(\menv,\msto,\mkont) .
\compile{\mexpi0}^\delta(\menv,\msto',\kif[^\mcntr]{\compile{\mexpi1},\compile{\mexpi2},\menv,\maddr})
\\
&\text{ where }\maddr,\msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)
\end{align*}
\caption{Compilation}
\end{figure}

@@

\appendix
\section{Relation to Uniform \(k\)-CFA (A Case Against Acceptability)}
\label{sec:accept}

\cite{dvanhorn:nielson-nielson-popl97} \cite{dvanhorn:Neilson:1999}

This machine's allocation strategy mimics the Uniform k-CFA analysis
in Principles of Program Analysis, which is defined in terms of
``$\delta$ contours.''  However, because the machine represetation makes
context explicit via continuations, we can calculate these contours
rather than thread them throught the evaluator.  In other words, we
can use the CESK* machine without modification to obtain Uniform k-CFA
by way of a simple allocation strategy.  (In this way, it's a
simplification of the presentation in JFP.)

NNH uses a coinductive acceptability relation to specify Uniform
k-CFA:

\[
   C,R \models^{ce}_\delta E
\]

The cache and global environment form a finite store-like structure
holding bindings and return values.  The contour environment ce maps
variables to locations in R which contains their bindings, just as the
environment of the CESK* machine does.  The current contour delta is a
string of application labels describing the enclosing context under
which this term is being analyzed (or evaluated).  If you view the
acceptability relation as a big-step evalator, the
$(\widehat C,\widehat\rho)$ component should be seen as a global
store ce is the environment mapping variables to their locations.

Starting form the initial configuration for a program and iterating
the machine transition relation until reaching a fixpoint of reachable
states will \emph{underestimate} the acceptability relation of Uniform
k-CFA.  You can recover acceptability by feeding this store back into
the initial configuration and iterating again.  Repeating this process
until a complete run of the program reaches no new states will be the
least solution that is acceptable.

HOWEVER.  Why should we care about acceptability?  What this
machine computes is safe.  In other words, it computes a more
precise characterization of the run-time behavior of a program.  In
doing is so, it actually saves work (as can be seen above).

An Example:

\begin{alltt}
 (let ((id (\(\lambda\) (x) x)))
   (begin (id 1) (id 2)))
\end{alltt}

Under Uniform 0-CFA, we would have:
\[
   [{\tt x} \mapsto \{{\tt 1}, {\tt 2}\}] \in \widehat\rho
\]

in the least solution to $\models$.  This says that, when run, 'x' is
bound to 1 or 2.

Under the machine semantics using a 0CFA allocation policy, the trace
semantics of the machine show that x is bound to 1, and that at some
later point, x becomes bound to 1 or 2.  Moreover, the machine would
show that (id 1) evaluates to 1 and only 1, while Uniform 0CFA must
give that (id 1) is either 1 or 2 to be acceptable.  We don't see any
value in these kinds of false flows that are due to the global and
timeless aspects of C,R which acceptability requires the heap to be
both finite and unchanging over the course of abstract
interpretation. (Another view of the difference: the machine abstracts
a program's execution as a \emph{finite state machine} that mimics the
machine interpretation of the program; the aceptability relation of
Uniform \(k\)-CFA abstracts a program's execution as a \emph{finite
  map} that mimics the big-step evaluator: from terms to (sets of)
values.)

\subsection{Another problem with acceptability: Temporal ignorance}

The small-step approach to static analysis brings subtle yet important temporal
richness not found in classical analyses for higher-order programs.
%
Classical analyses (ultimately) compute judgments on program terms and
contexts, e.g., at expression $e$, value $x$ may have value $v$.
%
The judgments do not relate the order in which expressions and context may be
evaluated in a program, e.g., a classical analysis has nothing to say with
regard to question like, ``Do we always evaluate $e_1$ before $e_2$?'' or ``Is
it always the case that a file handle is opened, read and then closed in that
order?''

Small-step analyses, by their nature, encode the temporal relationships between
abstract states.
%
It is sensible to make temporal queries of a small-step analysis.
%
Of course, this does not come for free: respecting temporal order imposes an
order in which states and terms may be evaluated \emph{during} the analysis.
%
Classical analyses can (and do) evaluate expressions in any order, or in some
cases, even in parallel~\cite{might:Prabhu:2010:EigenCFA}.
%
Relaxing that restriction on order affords additional optimizations that we
have \emph{not} performed.

We avoid sacrificing order not simply because we are interested in the
questions it allows us to ask, but because considering temporal order actually
improves the precision of the analysis itself.

\section{Pushdown Analysis}

It is straightforward to instantiate a \emph{pushdown} abstraction by
bounding only the variable binding portion of the heap, but using a
unique allocation strategy for continuations.  Such a strategy
abstracts a program's execution as a \emph{pushdown automata}
that mimics the machine interpretation.  This strategy therefore
models the abstract stack in a true stack like fashion and always
properly matches function calls with their return.

Although such analyses can be formulated straightforwardly in the
abstract machine approach, it is not clear all of the techniques of
this paper can be applied to similar effect in the pushdown context.
The main problem is calculation of an analysis can no longer be
computed as the fixed point of the machine transition relation.
Although there are several implementations (CFA2,ICFP'12), they
operate at speeds roughly on par with our starting point: unoptimized
store widened
machines. \cite{dvanhorn:Earl2012Introspective,dvanhorn:Vardoulakis2011CFA2}

>>>>

\paragraph{Primitives}

\begin{align*}
\mnum+1 &\in \interpdelta(\saddone,\mnum) &
\mnum-1 &\in \interpdelta(\ssubone,\mnum)\\
\strue &\in \interpdelta(\szerohuh,\szero) &
\sfalse &\in \interpdelta(\szerohuh,\mnum)\text{ if }\mnum\neq \szero\\
\end{align*}

\begin{align*}
\sNum &\in \hat\interpdelta(\saddone,\mnum) &
\sNum &\in \hat\interpdelta(\ssubone,\mnum)\\
\strue &\in \hat\interpdelta(\szerohuh,\sNum) &
\sfalse &\in \hat\interpdelta(\szerohuh,\sNum)\\
\strue &\in \hat\interpdelta(\szerohuh,\szero) &
\sfalse &\in \hat\interpdelta(\szerohuh,\mnum)\text{ if }\mnum\neq \szero\\
\end{align*}

%% \begin{figure}
%% \begin{alltt}
%%   ;; State \(\to\) \(\Set(\)State\()\)
%%   (define (step \(\mstate\))
%%     (match \(\mstate\)
%%       [(ev \(\mexp\) \(\rho\) \(\msto\) \(\mkont\))
%%        (match \(\mexp\)
%%          [(var\(\superscript\mlab\) \(\mvar\))
%%           (for/set ((\(\mval\) (lookup \(\rho\) \(\msto\) \(\mvar\))))
%%             (co \(\msto\) \(\mkont\) \(\mval\)))]
%%          [(lit\(\superscript\mlab\) \(\mlit\)) (set (co \(\msto\) \(\mkont\) \(\mlit\)))]
%%          [(lam\(\superscript\mlab\) \(\mvar\) \(\mexp\)) (set (co \(\msto\) \(\mkont\) (clos \(\mvar\) \(\mexp\) \(\rho\))))]
%%          [(app\(\superscript\mlab\) \(\mexp\subscript0\) \(\mexp\subscript1\))
%%           (define-values (\(\msto'\) \(\maddr\)) (push \(\mstate\)))
%%           (set (ev \(\mexp\subscript0\) \(\rho\) \(\msto'\) (ar \(\mexp\subscript1\) \(\rho\) \(\maddr\))))]
%%          [(ife\(\superscript\mlab\) \(\mexp\subscript0\) \(\mexp\subscript1\) \(\mexp\subscript2\))
%%           (define-values (\(\msto'\) \(\maddr\)) (push \(\mstate\)))
%%           (set (ev \(\mexp\subscript0\) \(\rho\) \(\msto'\) (ifk \(\mexp\subscript1\) \(\mexp\subscript2\) \(\rho\) \(\maddr\))))])]
%%       [(co \(\msto\) \(\mkont\) \(\mval\))
%%        (match \(\mkont\)
%%          ['mt (set (ans \(\msto\) \(\mval\)))]
%%          [(ar\(\superscript\mlab\) \(\mexp\) \(\rho\)) (set (ev \(\mexp\) \(\rho\) \(\msto\) (fn \(\mval\) l)))]
%%          [(fn\(\superscript\mlab\) \(\mval'\))
%%           (for/set ((\(\mkont\) (get-cont \(\msto\) l)))
%%             (ap \(\msto\) \(\mval'\) \(\mval\) \(\mkont\)))]
%%          [(fi\(\superscript\mlab\) c a \(\rho\))
%%           (for/set ((k (get-cont \(\msto\) l)))
%%             (ev (if v c \(\maddr\)) \(\rho\) \(\msto\) \(\mkont\)))])]
%%       [(ap \(\msto\) fun \(\maddr\) \(\mkont\))
%%        (match fun
%%          [(clos l \(\mvar\) \(\mexp\) \(\rho\))
%%           (define-values (\(\rho'\) \(\msto'\)) (bind \(\mstate\)))
%%           (set (ev \(\msto'\) \(\mexp\) \(\rho'\) \(\mkont\)))]
%%          [(op \(\mop\))
%%           (for*/set ((\(\mkont\) (get-cont \(\msto\) l))
%%                      (\(\mval\) (\(\interpdelta\) \(\mop\) \(\mval\))))
%%             (co \(\msto\) \(\mkont\) \(\mval\)))]
%%          [_ (set)]))]))
%% \end{alltt}
%% \caption{Implementation of machine transition relation.}
%% \end{figure}

\newpage
\section{Baseline analyzer}
\label{sec:baseline}

@@

\section{Implementation techniques}
\label{sec:opt}

@@

\begin{itemize}
 \item{[SS] quotient by sets of values (lazy non-determinism)}
 \item{[SS] remove {\tt ev} states (abstract compilation)}
 \item{[DS] operate on store changes rather than whole stores (store deltas)}
 \item{[DS] represent the store as a global, mutable hash and
   determine a state's store by the number of changes made to the
   global store at the time of visiting (timestamp)}
 \item{[DS] use a global mutable workset}
 \item{[DS] represent the global store as a global, mutable vector in
   the 0-CFA case. (preallocation)}
 \item{[SS] coarsely abstract literal data (abstracting compound data)}
\end{itemize}

Some techniques preserve the precision of the underlying analysis, and
others do not. We will discuss the design decisions behind the
precision-lossy techniques.  Each technique requires minimal changes
to the reduction relation/fixpoint, and each improves performance by
integer factors. The first three remain in the purely functional
realm. The final three use mutation in small, localized places, so
they do not increase conceptual complexity drastically.

@@

% Lazy:  cpu time: 32481 real time: 32881 gc time: 547

@@

The compilation step converts expressions into functions that expect
the other components of the {\tt ev} state. Its definition in figure
\ref{fig:compile} shows close similarity to the rules for interpreting
    {\tt ev} states. The next step is to change creation of {\tt ev} states to simply calling these functions. The modified reduction relation is shown in figure \ref{fig:caam}.

\begin{figure}
\begin{align*}
\compile{\svar\mvar} &= \lambda(\menv,\msto,\mkont) .\co{\mkont,\mval,\msto} \text{ where }\mval\in\msto(\menv(\mvar))
\\
\compile{\slit\mlit} &= \lambda(\menv,\msto,\mkont) .
\co{\mkont,\mlit,\msto}
\\
\compile{\slam\mvar\mexp} &= \lambda(\menv,\msto,\mkont) .
\co{\mkont,\clos{\mvar,\compile\mexp,\menv},\msto}
\\
\compile{\sapp[^\mlab]{\mexpi0}{\mexpi1}} &= \lambda^\mcntr(\menv,\msto,\mkont) .
\compile{\mexpi0}^\mcntr(\menv,\msto',\kar[_\mlab^\mcntr]{\compile{\mexpi1},\menv,\maddr})
\\
&
\text{ where }\maddr,\msto' = \mathit{push}^\mcntr_\mlab(\msto,\mkont)
\\
\compile{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2}} &= \lambda^\mcntr(\menv,\msto,\mkont) .
\compile{\mexpi0}^\delta(\menv,\msto',\kif[^\mcntr]{\compile{\mexpi1},\compile{\mexpi2},\menv,\maddr})
\\
&\text{ where }\maddr,\msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)
\end{align*}
\caption{Compilation}
\label{fig:compile}
\end{figure}

@@

% \include{appendix}
